Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models

### 2020-06-16 analysis/paper/paper.Rmd
incorporated comments from Richie, and the new references he added.

  Establishing metrics of quality habitat is critical for the recovery of endangered Pacific salmonids. A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but rely on linear fish-habitat relationships and are limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. QRF models allow for the consideration of noisy data, correlated variables, non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout the Columbia River Basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
keywords: |
  carrying capacity; quantile regression; random forests; Chinook; quantile regression forest
highlights: |
  These are the highlights. 
---

@@

@Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal. Since that review, there has been progress in identifying some fish-habitat relationships for some salmonid species. @Nickelson1992 demonstrated that juvenile coho were found in higher densities within pool habitat on the Oregon coast. Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001]. @Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools. @Braun2011 similarly found positive associations between spawner densities of sockeye in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools. Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002]. @Mossop2006 examined juvenile Chinook in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient. These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed. In addition, they all assumed some form of linear fish-habitat relationship which often results in weak predictive power.

A number of studies have utilized other modeling approaches to elicit non-salmonid fish habitat relationships. @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout densities and the width:depth ratio of a stream for the upper quantiles of trout density. The same approach was also used to map the potential extent of sole in the English Channel and southern North Sea [@Eastwood2003]. Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012]. @Wall2016 took a bioenergic approach to predict the capacity of a reach to support juvenile steelhead. The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships.

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Sweka2010 estimated carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners which is not usually available [@Cramer2009].

In fisheries it has long been recognized that both biotic and abiotic factors limit productivity within and across life-stages. For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage. We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality. Furthermore, we assert that observed fish density is a poor predictor of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity. To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) [@Meinshausen2006] models using measurements of fish abundance (and density) and habitat characteristics. QRF models combine the theory and justification of quantile regression modeling [@Koenker1978; @Cade2003] with the flexibility and framework of random forest models [@Breiman2001]. They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density. Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts [@Knudby2010]. Quantile regression forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers [@Prasad2006]. Meanwhile, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].

@@

CHaMP sites are 200 m to 500 m reaches within wadeable streams across select basins within the interior Columbia River basin selected based on a spatially balanced Generalized Random Tesselation Stratified sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, stream size (depth, width, discharge), substrate, temperature, and water quality.

@@

To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of Generalized Random Tesselation Stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two extrapolation models to each type of fish density, one that included watershed as covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear and areal capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart. Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).

To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists. The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.

@@

@@

We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m). We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average August temperature, wetted width:depth ratio, percent fines less than 6 mm, total percent fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).

@@

We fit a linear regression extrapolation using QRF-based predictions of capacity at all CHaMP sites as the response, and various GAAs as the independent variables. The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars). From this, we have estimates of capacity at every master sample point in the Columbia River basin, which are located approximately one kilometer apart.

@@

Estimates of Chinook salmon parr capacity from the QRF and extrapolation models were comparable to independent estimates from spawner-recruit data (Table \@ref(tab:sr-table), Figure \@ref(fig:sr-figure)). QRF estimates had overlapping confidence intervals with one or more of the Beverton-Holt, Ricker, or hockey stick model estimates in each of the `r as.character(english(length(keep_wtsds)))` locations where comparisons were possible (Figure \@ref(fig:sr-figure)). Possible additional uncertainty was not accounted for in estimates of spawners-per-redd or spawners-per-parr, which would increase the confidence intervals around spawner-recruit estimates and the overlap among estimates. Correlations between parr capacity estimates from the QRF model and spawner-recruit models ranged from 0.710 (Beverton-Holt) to 0.966 (Ricker). This favorable comparison provides strong validation as the spawner-recruit estimates of Chinook salmon parr capacity were developed from completely independent datasets and using entirely different methods.

@@

In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months in wadeable streams. Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act delisting. The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations. Moreover, these models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult). Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stage specific habitat patches may be limiting. As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support the target abundance. In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat. Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions.

The favorable comparison between QRF estimates of watershed capacity and the spawner-recruit estimates in selected watersheds goes a long way to justifying this approach. Although built from completely different data, when these mulitple lines of evidence converge it lends credence to the QRF capacity prediction results.

@@

Historically, fish-habitat relationship models have focused on species distribution or average abundance/density rather than directly addressing carrying capacity. Bioenergics approaches, such as the net rate of energy intake (NREI) have been applied to salmonids to estimate capacity on the same reach scale that our QRF model operates on [@Wall2016]. However, there are some potential issues with how the food supply (i.e. invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016]. The QRF approach described here allows us to analyze multiple (often correlated) habitat metrics and assess the potentially non-linear fish habitat relationships that define capacity as function of habitat.

@@

The biggest driver of capacity identified in this study is stream size, whether measured by wetted width or discharge, which should be unsurprising since we are using fish per meter as our response. In many ways, these metrics define habitat quantity; however, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sediment, and higher channel unit frequencies (a measure of habitat complexity and another way of saying more pool-riffle interfaces) and fish cover.

@@

Fish are mobile creatures and determining the appropriate spatial scale to estimate how their capacity may be determined by habitat characteristics is important. In the summer, for Chinook parr, there is clearly movement between multiple channel units (e.g. pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit. However, it is unlikely that they are moving up and down the entire watershed and we believe the 200 - 500 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity. At the same time, we acknowledge that managers, life-cycle modelers, and others are often interested in capacity estimates at larger spatial scales. While our QRF model can provide site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for estimates at larger spatial extents, such as watershed and population level scales. This is an efficient technique to leverage existing relationships for meaningful management decisions.

@@

One of the potential downsides to the extrapolation approach used here is that the globally available attributes generally do not change through time, and therefore may not reflect the dynamic nature of changing stream habitat. While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model. This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale.

@@

Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the Columbia River basin, there is a pressing need to develop new tools for habitat analyses. Unmanned Aerial Systems (UAS or drone, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015]. UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling. Advances in imaging techniques (e.g., multispectral imaging) and post-processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2018].

Development of a standardized protocol to incorporate remotely sensed data (LiDAR, aerial imagery) into the collection of habitat metrics would greatly increase the broadscale application of QRF. Rapid advances in drone technology further improve upon traditional habitat data collection by leveraging 1) sub-meter global navigation satellite system (GNSS) receivers; 2) cost-effective drone imagery collection, image stitching, and photogrammetry; and 3) semi-automated data post-processing. All data collection efforts would be georeferenced and topologically compatible to increase repeatability of methods and data collection locations; a primary criticism of previous CHaMP survey efforts. The implementation of such a protocol would circumvent the need to extrapolate by collecting data for individual channel units in a rapid manner using remote sensing technologies, thereby reducing labor, providing a cost-effective tool for habitat data collection supporting status and trend evaluation and model products to better inform habitat rehabilitation prioritization and planning.

@@

If a species' carrying capacity is defined or constrained, at least in part, by the habitat in which it lives, then illuminating statistically how such habitat impacts carrying capacity can lead to understanding how a species interacts with its environment. This understanding could be of crucial importance in the realm of conservation when dealing with an endangered or threatened species, but species/habitat interactions are a core element of ecological studies more generally. We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict an habitat's capacity. As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys. The framework could also be applied to depleted, non-migratory, and isolated populations (e.g. desert pupfish *Cyprinodon macularius*) to identify limiting factors in populations and/or determine whether a given habitat patch could support a viable population if limiting factors were addressed. Capacity estimates could also be used to evaluate potential translocation sites to determine if those sites could support an abundance considered viable before investing in translocation efforts.

>>>>

  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids. A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but rely on linear fish-habitat relationships and are limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
keywords: |
  carrying capacity; quantile regression; random forests; Chinook; quantile regression forest
highlights: |
  These are the highlights. 
---

@@

@Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal. Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species. @Nickelson1992 demonstrated that juvenile coho were found in higher densities within pool habitat on the Oregon coast. Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001]. @Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools. @Braun2011 similarly found positive associations between spawner densities of sockeye in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools. Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002]. @Mossop2006 examined juvenile Chinook in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient. These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed. In addition, they all assumed some form of linear fish-habitat relationship which often results in weak predictive power.

A number of studies have utilized other modeling approaches to elicit non-salmonid fish habitat relationships. @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout densities and the width:depth ratio of a stream for the upper quantiles of trout density. The same approach was also used to map the potential extent of sole in the English Channel and southern North Sea [@Eastwood2003]. Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012]. @Wall2016 took a bioenergic approach to predict the capacity of a reach to support juvenile steelhead. The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships and provide motivation for furthering research in this realm..

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Sweka2010 estimated carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners, data which is usually unavailable [@Cramer2009].

<!-- Revise the first sentence of the following paragraph, it seems out of place with the following sentence and paragraph -->

**In fisheries it has long been recognized that both biotic and abiotic factors limit productivity within and across life-stages.** For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage. We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality. Furthermore, we assert that observed fish density is a poor predictor of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity. To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) [@Meinshausen2006] models using measurements of fish abundance (and density) and habitat characteristics. QRF models combine the theory and justification of quantile regression modeling [@Koenker1978; @Cade2003] with the flexibility and framework of random forest models [@Breiman2001]. They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density. Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts [@Knudby2010]. Quantile regression forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers [@Prasad2006]. Meanwhile, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].

@@

CHaMP sites are 200 m to 600 m reaches within wadeable streams across select watershed within the interior Columbia River basin. The sites were selected based on a spatially balanced Generalized Random Tesselation Stratified sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, stream size (depth, width, discharge), substrate, temperature, macroinvertebrate productivity, and water quality.

@@

To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear (parr / m) and areal (parr / m$^2$) capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of Generalized Random Tesselation Stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two extrapolation models to each type of fish density, one that included watershed as covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear and areal capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart. Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).

To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the master sample points and lengths of streams to those within the domain of spring/summer Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists. The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.

@@

All code and data for the analyses presented here can be found on a [GitHub repository](https://github.com/KevinSee/QRFpaper).

@@

We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate the Maximal Information Criteria (MIC) value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (parr/m) (See Appendix S1 for further details). We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average August temperature, wetted width:depth ratio, percent fines less than 6 mm, total percent fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).

@@

We fit a linear regression extrapolation using QRF-based predictions of capacity at all CHaMP sites as the response, and various GAAs as the independent variables. The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars). From this, we calculated estimates of capacity at every master sample point in the Columbia River basin, at roughly a one kilometer scale.

@@

Estimates of Chinook salmon parr capacity from the QRF and extrapolation models were comparable to independent estimates from spawner-recruit data (Table \@ref(tab:sr-table), Figure \@ref(fig:sr-figure)). QRF estimates had overlapping confidence intervals with one or more of the Beverton-Holt, Ricker, or hockey stick model estimates in each of the `r as.character(english(length(keep_wtsds)))` locations where comparisons were possible (Figure \@ref(fig:sr-figure)). Potential additional uncertainty was not accounted for in estimates of spawners-per-redd or spawners-per-parr, which would increase the confidence intervals around spawner-recruit estimates and the overlap among estimates. Correlations between parr capacity estimates from the QRF model and spawner-recruit models ranged from 0.710 (Beverton-Holt) to 0.966 (Ricker). This favorable comparison provides strong validation as the spawner-recruit estimates of Chinook salmon parr capacity were developed from completely independent datasets and using entirely different methods.

@@

In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months in wadeable streams. Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act delisting. The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations. Moreover, these models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult). Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stage specific habitat patches may be limiting. As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support the target abundance. In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat. Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions and help guide not only the type of action, but also the location at which an action is performed.

The favorable comparison between QRF estimates of watershed capacity and the spawner-recruit estimates in selected watersheds helps support and validate this approach. Although built from completely different data, when these multiple lines of evidence converge it lends credence to the QRF capacity prediction results.

@@

Historically, fish-habitat relationship models have focused on species distribution or average abundance/density rather than directly addressing carrying capacity. Bioenergics approaches, such as the net rate of energy intake (NREI) have been applied to salmonids to estimate capacity on the same reach scale that our QRF model operates on [@Wall2016]. However, there are some potential issues with how the food supply (i.e. invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016] as well as difficulty in properly constraining drift depletion and inter-species competition, and computational and spatial limitation of this modelling approach [@Carmichael2020; @Wall2016]. The QRF approach described here allows us to analyze multiple (often correlated) habitat metrics and assess the potentially non-linear fish habitat relationships that define capacity as function of habitat.

@@

The biggest driver of capacity identified in this study is stream size, whether measured by wetted width or discharge, which should be unsurprising since we are using fish per meter as our response. In many ways, these metrics define habitat quantity; however, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sediment, and higher channel unit frequencies (a measure of habitat complexity and surrogate for the number of pool-riffle sequences or potential sheer areas providing feeding zones), and fish cover.

<!-- Richie's comment on the paragraph below: I would say that this is an important part, but I question whether it is appropriate to include in the manuscript? I am fine either way. -->

@@

Fish are mobile creatures and determining the appropriate spatial scale to estimate how their capacity may be determined by habitat characteristics is important. In the summer, for Chinook parr, there is clearly movement between multiple channel units (e.g. pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit. However, it is unlikely that they are moving up and down the entire watershed and we believe the 200 - 500 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity. At the same time, we acknowledge that managers, life-cycle modelers, and others are often interested in capacity estimates at larger spatial scales. While our QRF model can provide site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for estimates at larger spatial extents, such as watershed and population levels. This is an efficient technique to leverage existing relationships for meaningful management decisions.

@@

One of the potential downsides to the extrapolation approach used here is that the globally available attributes generally do not change through time, and therefore may not reflect the dynamic nature of changing stream habitat. While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model. This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale, but with improvements in remote sensing, spatially continuous data (modeled or measured) maybe on the horizon [@Tonia2019] (Tonina et al. 2019).

@@

Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the Columbia River basin, there is a pressing need to develop new tools for habitat analyses. Unmanned Aerial Systems (UAS or drones, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015]. UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling. Advances in imaging techniques (e.g., multispectral imaging) and post-processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2018]. Further, developments in Light Detection and Ranging (LiDAR) technology have allowed for the characterization of watershed scale geomorphologic and hydraulic variables not previously possible [@Tonina2019; @McKean2008].

Development of a standardized protocol to incorporate remotely sensed data (LiDAR, aerial imagery) into the collection of habitat metrics would greatly increase the broadscale application of QRF. Rapid advances in drone technology further improve upon traditional habitat data collection by leveraging 1) sub-meter global navigation satellite system (GNSS) receivers; 2) cost-effective drone imagery collection, image stitching, and photogrammetry; and 3) semi-automated to automated data post-processing. All data collection efforts would be georeferenced and topologically compatible to increase repeatability of methods and data collection locations; a primary criticism of previous CHaMP survey efforts. The implementation of such a protocol would circumvent the need to extrapolate by collecting data for individual channel units in a rapid manner using remote sensing technologies, thereby reducing labor, providing a cost-effective tool for habitat data collection supporting status and trend evaluation and model products to better inform habitat rehabilitation prioritization and planning.

@@

If a species' carrying capacity is defined or constrained, at least in part, by the habitat in which it lives, then illuminating statistically how such habitat impacts carrying capacity can lead to understanding how a species interacts with its environment. This understanding could be of crucial importance in the realm of conservation when dealing with an endangered or threatened species, but species/habitat interactions are a core element of ecological studies more generally. We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict a habitat's capacity. As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys. The framework could also be applied to depleted, non-migratory, and isolated populations (e.g. desert pupfish *Cyprinodon macularius*) to identify limiting factors in populations and/or determine whether a given habitat patch could support a viable population if limiting factors were addressed. Capacity estimates could also be used to evaluate potential translocation sites to determine if those sites could support an abundance considered viable before investing in translocation efforts.
